---
title: "Handling of ODE-models as R-objects (*rodeo*)"
author: "David Kneis (david.kneis [at] tu-dresden.de)"
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
---

<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Using rodeo}
-->

## What the package is about
 The *rodeo* package 
TODO

## Example problem

TODO

## Basic use

### Creating and inspecting model objects

We start by creating a new object with `new()`. This requires us to supply the name of the class as well as a set of data frames for initialization.

```{r createObj}
library(rodeo, quietly=TRUE)

# Load sample data frames
data(identifiers, processes, stoichiometry)

# Instantiate new object
model= new("rodeo", vars=subset(identifiers,type=="v"),
  pars=subset(identifiers,type=="p"), funs=subset(identifiers,type=="f"),
  pros=processes, stoi=stoichiometry)
```

To inspect the object's contents, we can use the following:
```{r inspectObj, eval=FALSE}
# Built-in method
model$show()

# Show stoichiometry information as a matrix
print(model$stoichiometryMatrix())
```

### Defining functions and supplying data

In order to work with the object, we need to define functions that are reference in the model's mathematical expressions (i.e. in the process rates or the stoichiometry factors).

```{r setData}
O2sat= function(t) {14.652 - 0.41022*t + 0.007991*t^2 - 0.000077774*t^3}
ka= function(u, d) {(0.728*sqrt(u) - 0.317*u + 0.0372*u^2) / d / 86400}
monod= function(s,h) {s / (s + h)}
```

We also need to set the values of parameters and state variables (initial values). We first define both parameters and initial values as lists. These lists are then coerced into named vectors using the dedicated methods `arrangePars()` and `arrangeVars()`. Use of these methods guarantees that the vector elements appear in proper order. Proper order is essential when the vector elements are accessed *by position* rather than *by name*.

```{r setFuns}
pars= list(kd=5.78e-7, h_do=0.5, s_do_z=2.76, wind=1, depth=2,
 temp=20, q_in=1, q_ex=1, c_z_in=0.1, c_do_in=9.022)
vars= list(c_z=1, c_do=9.022, v=1.e6)
p= model$arrangePars(pars)
v= model$arrangeVars(vars)
```

### Plotting a model object

Having defined all functions and having set data, we can plot the stoichiometry matrix. This is very useful for complex models with many variables and processes.

<table border = 0>
<tr>
<td>
```{r plotObjCode, eval=FALSE}
model$plot(c(v, p))
```
</td>
<td>
```{r plotObjResult, ref.label="plotObjCode", echo=FALSE, fig.width=2, fig.height=2, fig.align="center"}
```
</td>
</tr>
</table>

### Translating the model into source code

In order to use the model for simulation, we need to transfer it into source code. This is also known as 'code generation'. Specifically, we want the code generator to create a function that returns the derivatives of the state variables with respect to time. In addition to the derivatives, the generated function also returns the values of all process rates (as diagnostic variables).

After generating the code, we need to make it executable. In R, we can use a combination of `eval()` and `parse()`. Alternatively, the generated code could be for loaded with `source()` after exporting it to a file (e.g. using `write()`). The latter method is needed if one wants to inspect (or even modify) the generated code.

```{r generateR}
code= model$generate(name="derivs",lang="r")
derivs= eval(parse(text=code))
```

### Solving the ODE system

Now we are ready to compute the evolution of the state variables over time by means of numerical integration. At this point, the proper order of the elements in the vectors `v` and `p` is essential since the generated function code uses *by-index* access! Therefore, `v` and `p` should always be set by calls to the `arrangeVars()` and `arrangePars()` method, respectively.

In addition to the dynamics of the state variables, we also get the dynamics of the process rates.

```{r solve-0d}
library(deSolve)
t= seq(0, 30*86400, 3600)
out= ode(y=v, times=t, func=derivs, parms=p, NLVL=1)
layout(matrix(1:8, ncol=4, byrow=TRUE))
plot(out, mfrow=NULL)
```

## Advanced topics

### Handling spatially distributed systems

A zero-dimensional case (single reactor) has been considered so far. We will now extend the model for a collection of reactors.

In the first step, we need to provide vectors (instead of scalars) for one variable and one parameter, at least. Variables and parameters with scalar values will be replicated to vectors of the required length.

```{r solve-1d, fig.width=8, fig.height=3}
nbox= 3
pars= list(kd=rep(5.78e-7, nbox), h_do=0.5, s_do_z=2.76, wind=1, depth=2,
  temp=20, q_in=1, q_ex=1, c_z_in=0.1, c_do_in=9.022)
vars= list(c_z=seq(from=0, to=50, length.out=nbox), c_do=9.022, v=1.e6)
p= model$arrangePars(pars)
v= model$arrangeVars(vars)

out= ode(y=v, times=t, func=derivs, parms=p, NLVL=nbox)
layout(matrix(1:nbox, nrow=1))
plot(out, which=paste0("c_do",1:nbox), mfrow=NULL)
```

### Definition of models in Fortran (instead of R)

Real-world models usually consist of many and lenghty mathematical expressions. Also, depending on the studied problem, the ODE solver may need to use (very) short time steps. Then, computation times become of serious concern. In those time-critical cases, it is recommended to generate source code for a fast, compilable language rather than for (slower) R. The compilable language currently supported by the package is *Fortran95*.

The next statement generates *Fortran95* code and exports it to a file for later compilation.
```{r generateF}
code= model$generate(name="derivs",lang="f95")
file_genCode= paste0(tempfile(), ".f95")
write(x=code, file=file_genCode)
cat("code written to",file_genCode)
```

Inspection of the generated *Fortran95* code reveals that the interface of the central subroutine "derivs" is

`subroutine derivs(time, var, par, NLVL, dydt, pro)`

However, according to the [vignette of the *deSolve* package](http://cran.r-project.org/web/packages/deSolve/vignettes/compiledCode.pdf) (page 6), the following interface is required

`subroutine derivs (neq, t, y, ydot, yout, ip)`

and an additional subroutine for parameter initialization (`initmod`) must to be supplied as well.

Consequently, we need to manually code in Fortran

  (1) a wrapper around the generated function to make its interface compatible with *deSolve*
  (2) the additional subroutine for parameter initialization
  (3) any functions used by the model

A suitable Fortan code for (1) and (2) is given below. Note that this code is fully generic except for the statement `NLVL=3` which defines the number of spatial boxes.

```{r fortranWrapper, echo=FALSE}
file_auxCode= "wrapperCode.f95"
text= readLines(file_auxCode, n=-1L, ok=TRUE, warn=TRUE, encoding="unknown", skipNul=FALSE)
text= paste(text,"\n")
cat(text)
```

The model-specific functions (3) could be defines as :

```{r fortranFunctions, echo=FALSE}
file_funCode= "functionsCode.f95"
text= readLines(file_funCode, n=-1L, ok=TRUE, warn=TRUE, encoding="unknown", skipNul=FALSE)
text= paste(text,"\n")
cat(text)
```

As a next step, we need to compile *(a)* the generated code and *(b)* all hand-written auxiliary code (boxes above) into a shared library. For that purpose, we make use of R's command line utility `R CMD SHLIB`. The 3 files passed to the command are the following:
  
  - *file_funCode* : Hand-coded source file with the model-specific function (box above)
  - *file_auxCode* : Hand-coded source file containing the wrapper and dimension specifications (box above)
  - *file_genCode* : File containing the generated source

```{r compileF, echo=TRUE}
dllname= "mySharedLib"
dllfile= paste0(dllname,.Platform$dynlib.ext)
command= paste0("R CMD SHLIB ",file_funCode," ",file_genCode," ",file_auxCode,
  " --preclean --clean -o ",dllfile)
if (system(command) != 0)
  stop(paste0("Error running '",command,"'"))
```

```{r, echo=FALSE}
# Clean up files from compilation
invisible(file.remove(list.files(pattern=".+[.]mod")))
```

We are now prepared to load the shared library and run the simulation based on the Fortran code. Note the additional arguments `dllname`, `initfunc`, and `nout` being passed to the numerical solver (use `?lsoda` to see the documentation for them). Note that setting a wrong value for `nout` easily makes R  crash.

```{r solve-1d-F, fig.width=8, fig.height=3}
dyn.load(dllfile)
out= ode(y=v, times=t, func="derivs_wrapped", parms=p, NLVL=nbox,
  dllname=dllname, initfunc="initmod", nout=nrow(model$PROS)*nbox)
layout(matrix(1:nbox, nrow=1))
dyn.unload(dllfile)
plot(out, which=paste0("c_do",1:nbox), mfrow=NULL)
```

```{r, echo=FALSE}
# Clean up dll file
invisible(file.remove(list.files(pattern=paste0(".+",.Platform$dynlib.ext))))
```
